Associative Data Structures for Website Filtering

### 2014-04-17 paper/paper.tex
typos

    \begin{abstract}
        We have researched the performance of the hash table, binary search tree, and skip list
        associative datastructures in the context of a web filter.  Our measurements were made with
        structures of various sizes (ranging from $2^8$ up to $2^{20}$) and with different possible
        key types (from 4 to 120 bytes).  In this paper, we describe our implementation and
        empirically show that given a sufficiently high number of entries, a hash table has the
        highest insertion and query speeds, while the skip list requires the least memory per
        element.
        % And when I say "hash table", I mean libstdc++'s unordered_map, and when I say bst I mean
        % libstdc++'s map, and if you took different implementations you'd probably get different
        % results...
    \end{abstract}

    \section{Introduction}
    A website filter is a program that categorises websites in order to restrict access to
    designated sites a user might visit. Commonly, the filter operates by keeping a blacklist of
    sites a user may not visit unless they have certain credentials, or alternatively, keeping a
    whitelist and prohibiting any sites not on it.  As the world-wide web is growing, these lists
    have grown larger to a point where the speed of the underlying data structure has become a key
    issue.  The time of both inserting and looking up websites must be kept minimal. Moreover, the
    increased number of entries make memory usage an equally significant concern.

@@

    While the decision of data structure might mostly be dependant on the number of entries, the
    complexity of the key might also be of concern. Where permissions per page would be preferred, a
    more resource optimal solution might be found in filtering only on domain or even IP adress.  In
    researching the influence of this we formulate a secondary research question:

    \begin{quotation}
        How does the complexity of the key influence the preformance of various associative data structures?
    \end{quotation}

@@

    After running a number of tests we chose to decrease the number of times we repeated each run.
    We had set out to perform each measurement 1000 times and take the average over that.  In
    parctice, however, the results obtained by performing only 100 runs did not vary greatly, and
    allowed us to perform more individual measurements.

@@

    \subsection{Difference in performance between different keytypes}

@@

    A similar issue occurs around 4000 or 8000 elements (depending on data structure).  For all
    three datastructures, the cost of inserting and querying grows at a higher rate than before.
    This is difficult to quantify, but it appears some extra load is added at that point.  We suspect it is related
    either to the allocator having to acquire more memory or the CPU cache running out and more
    cache misses occuring.

@@

    Our secondary point of research encompassed the effect of key complexity on associative data
    structures.  Our different types of keys were IP adresses, domain names and full path URLs.  Our
    measurements confirmed our intuition that IP addresses impose significantly lower costs over the
    other two key types, as well as requiring less memory.  This improvement was more noticable in
    the binary search tree than in the two other systems.  The corresponding expected improvement
    when comparing the domain key type to the full path key type were not observed.

@@

    Knowing the strengths and weaknesses of the various data structures, further research could
    compare subvariants of the hash table and skip list.  In particular, it may be of interest to
    experiment with differen choices of hash function and collision resolution scheme.  Larger
    numbers of elements may also provide interesting challenges, as the number of elements may be
    too great to keep in main memory.  Finally, the situation where a map is accessed by multiple
    threads concurrently provides extra complications that could be considered.

>>>>

    \begin{abstract}
        We have researched the performance of the hash table, binary search tree, and skip list
        associative data structures in the context of a web filter.  Our measurements were made with
        structures of various sizes (ranging from $2^8$ up to $2^{20}$) and with different possible
        key types (from 4 to 120 bytes).  In this paper, we describe our implementation and
        empirically show that given a sufficiently high number of entries, a hash table has the
        highest insertion and query speeds, while the skip list requires the least memory per
        element.
        % And when I say "hash table", I mean libstdc++'s unordered_map, and when I say bst I mean
        % libstdc++'s map, and if you took different implementations you'd probably get different
        % results...
    \end{abstract}

    \section{Introduction}
    A website filter is a program that categorises websites in order to restrict access to
    designated sites a user might visit. Commonly, the filter operates by keeping a blacklist of
    sites a user may not visit unless they have certain credentials, or alternatively, keeping a
    whitelist and prohibiting any sites not on it.  As the world wide web is growing, these lists
    have grown larger to a point where the speed of the underlying data structure has become a key
    issue.  The time of both inserting and looking up websites must be kept minimal. Moreover, the
    increased number of entries make memory usage an equally significant concern.

@@

    While the decision of data structure might mostly be dependent on the number of entries, the
    complexity of the key might also be of concern. Where permissions per page would be preferred, a
    more resource optimal solution might be found in filtering only on domain or even IP address.  In
    researching the influence of this we formulate a secondary research question:

    \begin{quotation}
        How does the complexity of the key influence the performance of various associative data structures?
    \end{quotation}

@@

    After running a number of tests we chose to decrease the number of times we repeated each run.
    We had set out to perform each measurement 1000 times and take the average over that.  In
    practice, however, the results obtained by performing only 100 runs did not vary greatly, and
    allowed us to perform more individual measurements.

@@

    \subsection{Difference in performance between different key types}

@@

    A similar issue occurs around 4000 or 8000 elements (depending on data structure).  For all
    three data structures, the cost of inserting and querying grows at a higher rate than before.
    This is difficult to quantify, but it appears some extra load is added at that point.  We suspect it is related
    either to the allocator having to acquire more memory or the CPU cache running out and more
    cache misses occurring.

@@

    Our secondary point of research encompassed the effect of key complexity on associative data
    structures.  Our different types of keys were IP addresses, domain names and full path URLs.  Our
    measurements confirmed our intuition that IP addresses impose significantly lower costs over the
    other two key types, as well as requiring less memory.  This improvement was more noticeable in
    the binary search tree than in the two other systems.  The corresponding expected improvement
    when comparing the domain key type to the full path key type were not observed.

@@

    Knowing the strengths and weaknesses of the various data structures, further research could
    compare subvariants of the hash table and skip list.  In particular, it may be of interest to
    experiment with different choices of hash function and collision resolution scheme.  Larger
    numbers of elements may also provide interesting challenges, as the number of elements may be
    too great to keep in main memory.  Finally, the situation where a map is accessed by multiple
    threads concurrently provides extra complications that could be considered.
